Open Clusters Characterization in Gaia DR2 Using ML Algorithms

### 2020-12-05 thesis/thesis.tex
thesis: Extend Method chapter

For the sake of simplicity, we are taking \emph{Melotte 22 (Messier 45)} \cite{elsanhoury2019ppmxl} to illustrate the whole method. This cluster is well studied
and their stars are not too mixed with other region stars.

@@

@@

\begin{itemize}
  \item \verb|regions| for storing cluster properties such as \emph{location}, \emph{diameter} and other properties
  \item \verb|gaiadr2_source| which contains data for all downloaded stars from Gaia
\end{itemize}

The second table is partitioned by region, storing all stars related to one cluster inside an independent table for optimized access performance.

Methods for retrieving information from the database are available through an instance of \verb|cdalvaro.DB|. This class contains methods
for recovering stars information for a given region as a Pandas \verb|DataFrame| ready for data analysis.

These containers are orchestrated with a \verb|docker-compose.yml| file which automatically launches one instance of each container.

The developed code for this work can be found inside the \verb|src| directory at GitHub:
\href{https://github.com/cdalvaro/machine-learning-master-thesis}{cdalvaro/machine-learning-master-thesis} repository.

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/melotte_22/features_melotte_22.png}
  \end{subfigure}
  \caption{Melotte 22 feature pairplot}
  \label{fig:features_melotte_22}
\end{figure}

So, in summary, these are the features we are going to use as sources for our clustering models:

@@

The following example can be found in the Jupyter notebook: \verb|cluster_characterization.ipynb| (available at \verb|src/notebooks|).
It shows how to estimate the best number of clusters for Melotte 22 by using \verb|cdalvaro.ml.estimate_n_clusters| method.

\begin{minted}{python}
  n_clusters, kmeans = estimate_n_clusters(melotte22_df, min_clusters=3,
                                           max_clusters=10, verbose=True)

  # Output
  # Silhouette score for 3 clusters: 0.5420
  # Silhouette score for 4 clusters: 0.5393
  # Silhouette score for 5 clusters: 0.5608
  # Silhouette score for 6 clusters: 0.5336
  # Silhouette score for 7 clusters: 0.5306
  # Silhouette score for 8 clusters: 0.4978
  # Silhouette score for 9 clusters: 0.5214
  # Silhouette score for 10 clusters: 0.5199
  # Best silhouette score is 0.5608 for 5 clusters
\end{minted}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/melotte_22/dec_hr_diagram_melotte_22.png}
  \end{subfigure}
  \caption{Melotte 22 H-R diagram with DEC characterization}
  \label{fig:dec_hr_diagram_melotte_22}
\end{figure}

>>>>

For the sake of simplicity, we are taking \emph{Melotte 22 (Messier 45)} \cite{elsanhoury2019ppmxl} to illustrate the whole method. This cluster is well studied
and their stars are not too mixed with other region stars, so images look clearly.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/services_diagram.pdf}
  \caption{Data retrieval and analysis}
  \label{fig:services_diagram}
\end{figure}

These containers are orchestrated with a \verb|docker-compose.yml| file which automatically launches one instance of each container.
Figure \ref{fig:services_diagram} shows a diagram describing the services architecture for this work.

@@

\begin{itemize}
  \item \verb|public.regions| for storing cluster properties such as \emph{location}, \emph{diameter} and other properties
  \item \verb|public.gaiadr2_source| which contains data for all downloaded stars from Gaia
\end{itemize}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/customdb_diagram.pdf}
  \caption{Custom DB diagram}
  \label{fig:customdb_diagram}
\end{figure}

The second table is partitioned by region, storing all stars related to one region inside an independent table for optimized access performance.
The partition tables are located inside \verb|gaiadr2| schema. See Figure \ref{fig:customdb_diagram}.

Although \verb|public.gaiadr2_source| is partitioned, there is no need for us to directly access to one of those tables.
We only need to know the \verb|id| for the desired region and query for its stars to the main table as follows:

\begin{minted}{sql}
  -- Get region id
  SELECT id FROM public.regions WHERE name = 'Melotte 22';
  -- output: id -> 302

  -- Get stars for the given id
  -- NOTE: PostgreSQL cannot optimize this query when region_id is
  --       provided as a subquery. This is why we get first the region id.
  SELECT * FROM public.gaiadr2_source WHERE region_id = 302;
\end{minted}

PostgreSQL knows that all stars we want to are located inside \verb|gaiadr2.region_302| so it does not need to query any other tables.
This allows us to download data for as many regions as we want to without loosing efficiency when accessing for that data.

\verb|gaiadr2_source| and its partitions contain all fields available at the Gaia DR2 database.
A complete list of these fields and their description can be found at the following link:
\sloppy\url{https://gea.esac.esa.int/archive/documentation/GDR2/Gaia_archive/chap_datamodel/sec_dm_main_tables/ssec_dm_gaia_source.html}

Table \ref{tab:important_fields} shows a brief description of the fields used in this work.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{r|c|l}
      \textbf{Name} & \textbf{Units} & \textbf{Description} \\
      \hline
      \verb|ra| & Angle [deg] & Right ascension \\
      \verb|dec| & Angle [deg] & Declination \\
      \verb|parallax| & Angle [mas] & Parallax \\
      \verb|pmra| & Angular Velocity [mas/year] & Proper motion in right ascension \\
      \verb|pmdec| & Angular Velocity [mas/year] & Proper motion in declination \\
      \verb|phot_g_mean_mag| & Magnitude [mag] & G-band mean magnitude \\
      \verb|bp_rp| & Magnitude [mag] & BP - RP colour \\
    \end{tabular}
    \caption{Properties and descriptions of the fields used in this work.}
    \label{tab:important_fields}
  \end{center}
\end{table}

The downloader takes data from the Gaia DR2 database and save the selection into our custom database without modifying any value,
so data preserves all its properties and units.

Python methods for retrieving information from the database are available through an instance of \verb|cdalvaro.DB|.
This class contains methods for recovering stars information for a given region as a Pandas \verb|DataFrame| ready for data analysis.

The developed code for this work can be found inside the \verb|src| directory at GitHub:
\href{https://github.com/cdalvaro/machine-learning-master-thesis}{cdalvaro/machine-learning-master-thesis} repository.
(For the printed version of this work, you can scan the QR code on cover for accessing to it.)

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/melotte_22/features_melotte_22.png}
  \end{subfigure}
  \caption{Melotte 22 feature pairplot}
  \label{fig:features_melotte_22}
\end{figure}

\pagebreak

In summary, these are the features we are going to use as sources for our clustering models:

@@

The following example is a snippet copied from \verb|cluster_characterization.ipynb| Jupyter notebook (available at \verb|src/notebooks|).
It shows how to estimate the best number of clusters for Melotte 22 by using \verb|cdalvaro.ml.estimate_n_clusters| method.

\begin{minted}{python}
  n_clusters, kmeans = estimate_n_clusters(melotte22_df,
                                           min_clusters=3, max_clusters=7)
  # Output
  # Silhouette score for 3 clusters: 0.5420
  # Silhouette score for 4 clusters: 0.5393
  # Silhouette score for 5 clusters: 0.5608
  # Silhouette score for 6 clusters: 0.5336
  # Silhouette score for 7 clusters: 0.5306
  # Best silhouette score is 0.5608 for 5 clusters
\end{minted}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/dec_hr_diagram_melotte_22.png}
      \caption{DEC H-R diagram}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/hr_diagram_melotte_22.png}
      \caption{Clusterix+TOPCAT H-R diagram}
    \end{subfigure}
  \end{subfigure}
  \caption{H-R diagram comparison between DEC clustering and Clusterix+TOPCAT method.}
  \label{fig:hr_diagram_comparison}
\end{figure}

Finally, Figure \ref{fig:hr_diagram_comparison} shows a comparison between the H-R diagram obtained by DEC clustering and the one obtained
using the Clusterix + TOPCAT method.
